---
title: "DataQualityDashboard Tutorial on a Single Tutorial DuckDB"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: auto
---

## Introduction

This tutorial uses a single tutorial database, `data/syntheaCDMWithDq.duckdb`, built from a committed raw Synthea export with curated data quality issues. The same DuckDB file contains both the transformed OMOP CDM schema, `cdm`, and the raw staging schema, `synthea_native`, so students can move directly from a DQD finding to the transformed row and then back to the upstream raw record.

The tutorial keeps the same overall teaching arc:

1. run DQD
2. inspect the failing check metadata
3. view the package SQL template behind the check
4. run a readable diagnostic query
5. trace the issue back to the raw staging schema when relevant
6. apply a local demonstration fix in a temporary working copy
7. rerun DQD to confirm the effect

## Environment and Package Setup

```{r setup}
library(DBI)
library(duckdb)
library(DatabaseConnector)
library(DataQualityDashboard)

repo_root_candidates <- c(
  getwd(),
  normalizePath(file.path(getwd(), ".."), winslash = "/", mustWork = FALSE)
)
repo_root <- repo_root_candidates[
  file.exists(file.path(repo_root_candidates, "R", "dqd_helpers.R"))
][1]

if (is.na(repo_root) || !nzchar(repo_root)) {
  stop("Could not locate the repository root from the Quarto working directory.")
}

source(file.path(repo_root, "R", "dqd_helpers.R"))
source(file.path(repo_root, "R", "tutorial_sql_helpers.R"))

tutorial_db <- file.path(repo_root, "data", "syntheaCDMWithDq.duckdb")
raw_issue_dir <- file.path(repo_root, "data", "syntheaRawWithDqIssues")

if (!file.exists(tutorial_db)) {
  stop("Missing tutorial database: ", tutorial_db, ". Run `Rscript BuildTutorialCdmWithDq.R` first.")
}
if (!dir.exists(raw_issue_dir)) {
  stop("Missing committed raw tutorial inputs: ", raw_issue_dir, ". Run `Rscript PrepareRawIssueSet.R` first.")
}

quarto_work_dir <- file.path(repo_root, "dqd_output", "quarto-work")
results_dir <- file.path(repo_root, "dqd_output", "results")
ensure_dir(quarto_work_dir)
ensure_dir(results_dir)

working_db <- file.path(quarto_work_dir, "tutorial-working.duckdb")
copy_database_file(tutorial_db, working_db, overwrite = TRUE)

connection_details <- DatabaseConnector::createConnectionDetails(
  dbms = "duckdb",
  server = working_db
)

prepare_dqd_support_tables(
  db_path = working_db,
  cdm_database_schema = tutorial_cdm_schema,
  results_database_schema = tutorial_meta_schema
)

tutorial_checks <- get_tutorial_check_metadata(cdm_version = "5.4")

show_sql_block <- function(sql_text) {
  cat("```sql\n", sql_text, "\n```\n", sep = "")
}

show_template_for_check <- function(sql_file) {
  show_sql_block(read_dqd_template(sql_file))
}
```

## Tutorial Assets

The key assets in this workflow are:

- clean raw source files in `data/syntheaRaw/`
- committed mutated raw source files in `data/syntheaRawWithDqIssues/`
- the single tutorial database `data/syntheaCDMWithDq.duckdb`

Inside `syntheaCDMWithDq.duckdb`:

- `cdm` holds the OMOP CDM tables DQD evaluates
- `synthea_native` holds the staged raw Synthea tables loaded by the ETL
- `main` holds tutorial metadata tables that track the curated issues and fixes

For the convention stage, the tutorial targets exactly two checks:

- `fkDomain` on `MEASUREMENT.MEASUREMENT_CONCEPT_ID`
- `isStandardValidConcept` on `DRUG_EXPOSURE.DRUG_CONCEPT_ID`

```{r tutorial-check-overview}
tutorial_checks[
  ,
  c(
    "target_check",
    "severity",
    "table_name",
    "field_name",
    "checkLevel",
    "sqlFile"
  )
]
```

## Stage 1: Fatal Checks

The two fatal tutorial targets are:

- `isRequired` on `VISIT_OCCURRENCE.VISIT_END_DATE`
- `measureObservationPeriodOverlap` on `OBSERVATION_PERIOD`

```{r fatal-run-dqd}
fatal_before_json <- file.path(results_dir, "quarto-single-fatal-before.json")
fatal_after_json <- file.path(results_dir, "quarto-single-fatal-after.json")

fatal_before <- DataQualityDashboard::executeDqChecks(
  connectionDetails = connection_details,
  cdmDatabaseSchema = tutorial_cdm_schema,
  resultsDatabaseSchema = tutorial_meta_schema,
  cdmSourceName = "SyntheaTutorialSingleFatal",
  numThreads = 1,
  sqlOnly = FALSE,
  outputFolder = results_dir,
  outputFile = basename(fatal_before_json),
  verboseMode = TRUE,
  writeToTable = FALSE,
  checkLevels = c("TABLE", "FIELD", "CONCEPT"),
  checkSeverity = "fatal",
  cdmVersion = "5.4"
)
```

```{r fatal-inspect-results}
fatal_before$CheckResults[
  fatal_before$CheckResults$checkName %in% c("isRequired", "measureObservationPeriodOverlap") &
    fatal_before$CheckResults$numViolatedRows > 0,
  c("checkName", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows", "sqlFile")
]
```

```{r fatal-template-required, results='asis'}
show_template_for_check(
  subset(tutorial_checks, target_check == "isRequired" & table_name == "VISIT_OCCURRENCE")$sqlFile[1]
)
```

```{r fatal-diagnostic-required, results='asis'}
show_sql_block(get_issue_sql("fatal_visit_end_date_required"))
```

```{r fatal-find-required}
fatal_connection <- connect_duckdb(working_db, read_only = TRUE)
missing_visit_end_dates <- dbGetQuery(fatal_connection, get_issue_sql("fatal_visit_end_date_required"))
disconnect_duckdb(fatal_connection)

missing_visit_end_dates
```

```{r fatal-template-overlap, results='asis'}
show_template_for_check(
  subset(tutorial_checks, target_check == "measureObservationPeriodOverlap")$sqlFile[1]
)
```

```{r fatal-diagnostic-overlap, results='asis'}
show_sql_block(get_issue_sql("fatal_observation_period_overlap"))
```

```{r fatal-find-overlap}
fatal_connection <- connect_duckdb(working_db, read_only = TRUE)
observation_period_overlap <- dbGetQuery(fatal_connection, get_issue_sql("fatal_observation_period_overlap"))
disconnect_duckdb(fatal_connection)

observation_period_overlap
```

```{r fatal-fix}
fatal_connection <- connect_duckdb(working_db, read_only = FALSE)
apply_fatal_fixes(fatal_connection, stage = "quarto-fatal", cdm_schema = tutorial_cdm_schema)
disconnect_duckdb(fatal_connection)
```

```{r fatal-rerun}
fatal_after <- DataQualityDashboard::executeDqChecks(
  connectionDetails = connection_details,
  cdmDatabaseSchema = tutorial_cdm_schema,
  resultsDatabaseSchema = tutorial_meta_schema,
  cdmSourceName = "SyntheaTutorialSingleFatalFixed",
  numThreads = 1,
  sqlOnly = FALSE,
  outputFolder = results_dir,
  outputFile = basename(fatal_after_json),
  verboseMode = TRUE,
  writeToTable = FALSE,
  checkLevels = c("TABLE", "FIELD", "CONCEPT"),
  checkSeverity = "fatal",
  cdmVersion = "5.4"
)

fatal_after$CheckResults[
  fatal_after$CheckResults$checkName %in% c("isRequired", "measureObservationPeriodOverlap") &
    fatal_after$CheckResults$numViolatedRows > 0,
  c("checkName", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows")
]
```

## Stage 2: Convention Checks

Convention is now centered on cross-schema investigation. One issue is driven directly from the committed mutated raw files. The other is a narrow post-ETL tutorial adjustment that is still traceable back to marked raw rows.

```{r convention-run-dqd}
convention_before_json <- file.path(results_dir, "quarto-single-convention-before.json")
convention_after_json <- file.path(results_dir, "quarto-single-convention-after.json")

convention_before <- DataQualityDashboard::executeDqChecks(
  connectionDetails = connection_details,
  cdmDatabaseSchema = tutorial_cdm_schema,
  resultsDatabaseSchema = tutorial_meta_schema,
  cdmSourceName = "SyntheaTutorialSingleConvention",
  numThreads = 1,
  sqlOnly = FALSE,
  outputFolder = results_dir,
  outputFile = basename(convention_before_json),
  verboseMode = TRUE,
  writeToTable = FALSE,
  checkLevels = c("TABLE", "FIELD", "CONCEPT"),
  checkSeverity = "convention",
  cdmVersion = "5.4"
)
```

```{r convention-inspect-results}
convention_before$CheckResults[
  convention_before$CheckResults$checkName %in% c("fkDomain", "isStandardValidConcept") &
    convention_before$CheckResults$numViolatedRows > 0 &
    convention_before$CheckResults$cdmFieldName %in% c("MEASUREMENT_CONCEPT_ID", "DRUG_CONCEPT_ID"),
  c("checkName", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows", "sqlFile")
]
```

### Convention Check 1: `fkDomain` on `MEASUREMENT.MEASUREMENT_CONCEPT_ID`

This failure starts from marked rows in `data/syntheaRawWithDqIssues/observations.csv`. The raw marker identifies the visit in `synthea_native.observations`, and the tutorial then adjusts one measurement row from that same visit to an `Observation` concept so DQD flags `MEASUREMENT.MEASUREMENT_CONCEPT_ID` for `fkDomain`.

```{r convention-fkdomain-template, results='asis'}
show_template_for_check(
  subset(
    tutorial_checks,
    target_check == "fkDomain" & table_name == "MEASUREMENT"
  )$sqlFile[1]
)
```

```{r convention-fkdomain-diagnostic-sql, results='asis'}
show_sql_block(get_issue_sql("convention_measurement_fk_domain"))
```

```{r convention-fkdomain-find}
convention_connection <- connect_duckdb(working_db, read_only = TRUE)
measurement_domain_failures <- dbGetQuery(convention_connection, get_issue_sql("convention_measurement_fk_domain"))
disconnect_duckdb(convention_connection)

measurement_domain_failures
```

### Convention Check 2: `isStandardValidConcept` on `DRUG_EXPOSURE.DRUG_CONCEPT_ID`

For this tutorial, the raw medication rows are clearly marked in `synthea_native.medications`, but a small post-ETL tutorial adjustment is used to guarantee a non-standard concept in `DRUG_EXPOSURE.DRUG_CONCEPT_ID`. The lineage table in `main` keeps that relationship explicit.

```{r convention-standardvalid-template, results='asis'}
show_template_for_check(
  subset(
    tutorial_checks,
    target_check == "isStandardValidConcept" & table_name == "DRUG_EXPOSURE"
  )$sqlFile[1]
)
```

```{r convention-standardvalid-diagnostic-sql, results='asis'}
show_sql_block(get_issue_sql("convention_drug_is_standard_valid_concept"))
```

```{r convention-standardvalid-find}
convention_connection <- connect_duckdb(working_db, read_only = TRUE)
drug_standard_valid_failures <- dbGetQuery(
  convention_connection,
  get_issue_sql("convention_drug_is_standard_valid_concept")
)
disconnect_duckdb(convention_connection)

drug_standard_valid_failures
```

### Cross-Schema Traceback

The same DuckDB file lets you inspect the raw rows behind the convention issues.

```{r convention-raw-observations}
convention_connection <- connect_duckdb(working_db, read_only = TRUE)
raw_fk_rows <- dbGetQuery(
  convention_connection,
  "
  SELECT DATE, PATIENT, ENCOUNTER, CODE, DESCRIPTION
  FROM synthea_native.observations
  WHERE DESCRIPTION LIKE '%[tutorial fkDomain issue]%'
  ORDER BY DATE, PATIENT, ENCOUNTER
  "
)
disconnect_duckdb(convention_connection)

raw_fk_rows
```

```{r convention-raw-medications}
convention_connection <- connect_duckdb(working_db, read_only = TRUE)
raw_standard_rows <- dbGetQuery(
  convention_connection,
  "
  SELECT START, PATIENT, ENCOUNTER, CODE, DESCRIPTION
  FROM synthea_native.medications
  WHERE DESCRIPTION LIKE '%[tutorial standard validity issue]%'
  ORDER BY START, PATIENT, ENCOUNTER
  "
)
disconnect_duckdb(convention_connection)

raw_standard_rows
```

```{r convention-fix}
convention_connection <- connect_duckdb(working_db, read_only = FALSE)
apply_convention_fixes(convention_connection, stage = "quarto-convention", cdm_schema = tutorial_cdm_schema)
disconnect_duckdb(convention_connection)
```

```{r convention-rerun}
convention_after <- DataQualityDashboard::executeDqChecks(
  connectionDetails = connection_details,
  cdmDatabaseSchema = tutorial_cdm_schema,
  resultsDatabaseSchema = tutorial_meta_schema,
  cdmSourceName = "SyntheaTutorialSingleConventionFixed",
  numThreads = 1,
  sqlOnly = FALSE,
  outputFolder = results_dir,
  outputFile = basename(convention_after_json),
  verboseMode = TRUE,
  writeToTable = FALSE,
  checkLevels = c("TABLE", "FIELD", "CONCEPT"),
  checkSeverity = "convention",
  cdmVersion = "5.4"
)

convention_after$CheckResults[
  convention_after$CheckResults$checkName %in% c("fkDomain", "isStandardValidConcept") &
    convention_after$CheckResults$cdmFieldName %in% c("MEASUREMENT_CONCEPT_ID", "DRUG_CONCEPT_ID") &
    convention_after$CheckResults$numViolatedRows > 0,
  c("checkName", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows")
]
```

## Stage 3: Characterization Checks

The characterization targets remain:

- `withinVisitDates`
- `plausibleStartBeforeEnd`

```{r characterization-run-dqd}
characterization_before_json <- file.path(results_dir, "quarto-single-characterization-before.json")
characterization_after_json <- file.path(results_dir, "quarto-single-characterization-after.json")

characterization_before <- DataQualityDashboard::executeDqChecks(
  connectionDetails = connection_details,
  cdmDatabaseSchema = tutorial_cdm_schema,
  resultsDatabaseSchema = tutorial_meta_schema,
  cdmSourceName = "SyntheaTutorialSingleCharacterization",
  numThreads = 1,
  sqlOnly = FALSE,
  outputFolder = results_dir,
  outputFile = basename(characterization_before_json),
  verboseMode = TRUE,
  writeToTable = FALSE,
  checkLevels = c("TABLE", "FIELD", "CONCEPT"),
  checkSeverity = "characterization",
  cdmVersion = "5.4"
)
```

```{r characterization-inspect-results}
characterization_before$CheckResults[
  characterization_before$CheckResults$checkName %in% c("withinVisitDates", "plausibleStartBeforeEnd") &
    characterization_before$CheckResults$numViolatedRows > 0,
  c("checkName", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows", "sqlFile")
]
```

```{r characterization-withinvisit-diagnostic, results='asis'}
show_sql_block(get_issue_sql("characterization_observation_within_visit_dates"))
```

```{r characterization-withinvisit-find}
characterization_connection <- connect_duckdb(working_db, read_only = TRUE)
observation_window_failures <- dbGetQuery(
  characterization_connection,
  get_issue_sql("characterization_observation_within_visit_dates")
)
disconnect_duckdb(characterization_connection)

head(observation_window_failures, 20)
```

```{r characterization-startbeforeend-diagnostic, results='asis'}
show_sql_block(get_issue_sql("characterization_drug_start_before_end"))
```

```{r characterization-startbeforeend-find}
characterization_connection <- connect_duckdb(working_db, read_only = TRUE)
drug_date_failures <- dbGetQuery(
  characterization_connection,
  get_issue_sql("characterization_drug_start_before_end")
)
disconnect_duckdb(characterization_connection)

drug_date_failures
```

```{r characterization-fix}
characterization_connection <- connect_duckdb(working_db, read_only = FALSE)
apply_characterization_fixes(
  characterization_connection,
  stage = "quarto-characterization",
  cdm_schema = tutorial_cdm_schema
)
disconnect_duckdb(characterization_connection)
```

```{r characterization-rerun}
characterization_after <- DataQualityDashboard::executeDqChecks(
  connectionDetails = connection_details,
  cdmDatabaseSchema = tutorial_cdm_schema,
  resultsDatabaseSchema = tutorial_meta_schema,
  cdmSourceName = "SyntheaTutorialSingleCharacterizationFixed",
  numThreads = 1,
  sqlOnly = FALSE,
  outputFolder = results_dir,
  outputFile = basename(characterization_after_json),
  verboseMode = TRUE,
  writeToTable = FALSE,
  checkLevels = c("TABLE", "FIELD", "CONCEPT"),
  checkSeverity = "characterization",
  cdmVersion = "5.4"
)

characterization_after$CheckResults[
  characterization_after$CheckResults$checkName %in% c("withinVisitDates", "plausibleStartBeforeEnd") &
    characterization_after$CheckResults$numViolatedRows > 0,
  c("checkName", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows")
]
```

## Wrap-up

The single-DB workflow makes the teaching loop tighter:

- DQD evaluates `cdm`
- the same DuckDB file keeps the raw staging tables in `synthea_native`
- `main` stores tutorial lineage and fix metadata

That lets students move from a failed check to the transformed row, then to the staged raw record, without switching databases or relying on prebuilt severity snapshots.
