---
title: "DataQualityDashboard Tutorial on a Staged OMOP CDM"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: auto
---

## Introduction

This tutorial walks through a staged DQD workflow on a synthetic OMOP CDM built from Synthea. The tutorial data has six curated issues distributed across `fatal`, `convention`, and `characterization` severities.

Each stage follows the same pattern:

1. run DQD for the current severity
2. inspect the failures
3. query the CDM to find the bad rows
4. apply the fix
5. rerun DQD to verify the fix

The prepared snapshots keep each stage deterministic without mutating the clean baseline.

## Environment and Package Setup

```{r setup}
library(DBI)
library(duckdb)
library(DataQualityDashboard)
library(jsonlite)

repo_root_candidates <- c(getwd(), normalizePath(file.path(getwd(), ".."), winslash = "/", mustWork = FALSE))
repo_root <- repo_root_candidates[
  file.exists(file.path(repo_root_candidates, "R", "dqd_helpers.R"))
][1]

if (is.na(repo_root) || !nzchar(repo_root)) {
  stop("Could not locate the repository root from the Quarto working directory.")
}

source(file.path(repo_root, "R", "dqd_helpers.R"))
source(file.path(repo_root, "R", "tutorial_sql_helpers.R"))

snapshot_paths <- list(
  broken = file.path(repo_root, "data", "syntheaCDM_tutorial_broken.duckdb"),
  after_fatal = file.path(repo_root, "data", "syntheaCDM_tutorial_after_fatal.duckdb"),
  after_convention = file.path(repo_root, "data", "syntheaCDM_tutorial_after_convention.duckdb")
)

missing_snapshots <- snapshot_paths[!file.exists(unlist(snapshot_paths))]
if (length(missing_snapshots) > 0) {
  stop(
    "Missing tutorial snapshot(s): ",
    paste(unlist(missing_snapshots), collapse = ", "),
    ". Run `Rscript PrepareTutorialSnapshots.R` first."
  )
}

quarto_work_dir <- file.path("dqd_output", "quarto-work")
quarto_work_dir <- file.path(repo_root, "dqd_output", "quarto-work")
ensure_dir(quarto_work_dir)
ensure_dir(file.path(repo_root, "dqd_output", "results"))
```

## Tutorial Assets and Snapshot Overview

The tutorial uses these prepared databases:

- `data/syntheaCDM_tutorial_broken.duckdb`: contains all injected issues
- `data/syntheaCDM_tutorial_after_fatal.duckdb`: fatal issues already fixed
- `data/syntheaCDM_tutorial_after_convention.duckdb`: fatal and convention issues already fixed

::: {.callout-note}
Each stage works on a temporary copy of the corresponding snapshot. That lets you rerun chunks without corrupting the prepared reference databases.
:::

## Stage 1: Fatal Checks

We begin from the fully broken tutorial snapshot and run only `fatal` DQD checks.

```{r fatal-prepare-working-db}
fatal_work_db <- file.path(quarto_work_dir, "fatal-working.duckdb")
copy_database_file(snapshot_paths$broken, fatal_work_db, overwrite = TRUE)
fatal_before_json <- file.path(repo_root, "dqd_output", "results", "quarto-fatal-before.json")
fatal_after_json <- file.path(repo_root, "dqd_output", "results", "quarto-fatal-after.json")
```

```{r fatal-run-dqd}
fatal_before <- run_dqd(
  db_path = fatal_work_db,
  cdm_source_name = "SyntheaTutorialFatal",
  check_severity = "fatal",
  output_json_path = fatal_before_json,
  verbose_mode = TRUE,
  write_to_table = FALSE
)
```

```{r fatal-inspect-results}
fatal_targets <- filter_dqd_results(
  fatal_before,
  check_names = c("isPrimaryKey", "measureObservationPeriodOverlap")
)

fatal_targets[
  fatal_targets$numViolatedRows > 0,
  c("checkName", "checkLevel", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows")
]
```

::: {.callout-note}
What to notice: the staged fatal section is centered on duplicate `MEASUREMENT_ID` values and overlapping `OBSERVATION_PERIOD` rows for a single person.
:::

## Stage 1 Investigation SQL

### Duplicate `MEASUREMENT_ID`

```{r fatal-duplicate-sql, results='asis'}
cat("```sql\n", get_issue_sql("fatal_duplicate_measurement_pk"), "\n```", sep = "")
```

```{r fatal-find-duplicate-measurement-id}
fatal_connection <- connect_duckdb(fatal_work_db, read_only = TRUE)
duplicate_measurements <- dbGetQuery(
  fatal_connection,
  get_issue_sql("fatal_duplicate_measurement_pk")
)
disconnect_duckdb(fatal_connection)

duplicate_measurements
```

### Overlapping observation periods

```{r fatal-overlap-sql, results='asis'}
cat("```sql\n", get_issue_sql("fatal_observation_period_overlap"), "\n```", sep = "")
```

```{r fatal-find-overlap}
fatal_connection <- connect_duckdb(fatal_work_db, read_only = TRUE)
observation_period_overlap <- dbGetQuery(
  fatal_connection,
  get_issue_sql("fatal_observation_period_overlap")
)
disconnect_duckdb(fatal_connection)

observation_period_overlap
```

## Stage 1 Fixes

Now apply the fatal fixes directly to the working copy.

```{r fatal-fix-issues}
fatal_connection <- connect_duckdb(fatal_work_db, read_only = FALSE)
apply_fatal_fixes(fatal_connection, stage = "quarto-fatal")
disconnect_duckdb(fatal_connection)
```

## Stage 1 Rerun and Verification

```{r fatal-rerun-dqd}
fatal_after <- run_dqd(
  db_path = fatal_work_db,
  cdm_source_name = "SyntheaTutorialFatalFixed",
  check_severity = "fatal",
  output_json_path = fatal_after_json,
  verbose_mode = TRUE,
  write_to_table = FALSE
)
```

```{r fatal-verify}
fatal_after_targets <- filter_dqd_results(
  fatal_after,
  check_names = c("isPrimaryKey", "measureObservationPeriodOverlap")
)

fatal_after_targets[
  fatal_after_targets$numViolatedRows > 0,
  c("checkName", "checkLevel", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows")
]
```

::: {.callout-note}
Expected outcome: the injected fatal findings should be resolved after the rerun.
:::

## Stage 2: Convention Checks

The convention stage starts from the prepared snapshot where the fatal issues are already fixed.

```{r convention-prepare-working-db}
convention_work_db <- file.path(quarto_work_dir, "convention-working.duckdb")
copy_database_file(snapshot_paths$after_fatal, convention_work_db, overwrite = TRUE)
convention_before_json <- file.path(repo_root, "dqd_output", "results", "quarto-convention-before.json")
convention_after_json <- file.path(repo_root, "dqd_output", "results", "quarto-convention-after.json")
```

```{r convention-run-dqd}
convention_before <- run_dqd(
  db_path = convention_work_db,
  cdm_source_name = "SyntheaTutorialConvention",
  check_severity = "convention",
  output_json_path = convention_before_json,
  verbose_mode = TRUE,
  write_to_table = FALSE
)
```

```{r convention-inspect-results}
convention_targets <- filter_dqd_results(
  convention_before,
  check_names = c("measurePersonCompleteness", "measureConditionEraCompleteness")
)

convention_targets[
  convention_targets$numViolatedRows > 0,
  c("checkName", "checkLevel", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows", "thresholdValue")
]
```

## Stage 2 Investigation SQL

### Missing procedure coverage

```{r convention-procedure-sql, results='asis'}
cat("```sql\n", get_issue_sql("convention_procedure_person_completeness"), "\n```", sep = "")
```

```{r convention-find-procedure-gap}
convention_connection <- connect_duckdb(convention_work_db, read_only = TRUE)
procedure_gap <- dbGetQuery(
  convention_connection,
  get_issue_sql("convention_procedure_person_completeness")
)
disconnect_duckdb(convention_connection)

head(procedure_gap, 20)
```

### Missing condition eras

```{r convention-condition-era-sql, results='asis'}
cat("```sql\n", get_issue_sql("convention_condition_era_completeness"), "\n```", sep = "")
```

```{r convention-find-condition-era-gap}
convention_connection <- connect_duckdb(convention_work_db, read_only = TRUE)
condition_era_gap <- dbGetQuery(
  convention_connection,
  get_issue_sql("convention_condition_era_completeness")
)
disconnect_duckdb(convention_connection)

condition_era_gap
```

## Stage 2 Fixes

```{r convention-fix-issues}
convention_connection <- connect_duckdb(convention_work_db, read_only = FALSE)
apply_convention_fixes(convention_connection, stage = "quarto-convention")
disconnect_duckdb(convention_connection)
```

## Stage 2 Rerun and Verification

```{r convention-rerun-dqd}
convention_after <- run_dqd(
  db_path = convention_work_db,
  cdm_source_name = "SyntheaTutorialConventionFixed",
  check_severity = "convention",
  output_json_path = convention_after_json,
  verbose_mode = TRUE,
  write_to_table = FALSE
)
```

```{r convention-verify}
convention_after_targets <- filter_dqd_results(
  convention_after,
  check_names = c("measurePersonCompleteness", "measureConditionEraCompleteness")
)

convention_after_targets[
  convention_after_targets$numViolatedRows > 0,
  c("checkName", "checkLevel", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows", "thresholdValue")
]
```

::: {.callout-note}
Expected outcome: the procedure coverage gap and missing condition era gap should no longer appear after the convention rerun.
:::

## Stage 3: Characterization Checks

The characterization stage starts from the prepared snapshot where fatal and convention issues are already fixed.

```{r characterization-prepare-working-db}
characterization_work_db <- file.path(quarto_work_dir, "characterization-working.duckdb")
copy_database_file(snapshot_paths$after_convention, characterization_work_db, overwrite = TRUE)
characterization_before_json <- file.path(repo_root, "dqd_output", "results", "quarto-characterization-before.json")
characterization_after_json <- file.path(repo_root, "dqd_output", "results", "quarto-characterization-after.json")
```

```{r characterization-run-dqd}
characterization_before <- run_dqd(
  db_path = characterization_work_db,
  cdm_source_name = "SyntheaTutorialCharacterization",
  check_severity = "characterization",
  output_json_path = characterization_before_json,
  verbose_mode = TRUE,
  write_to_table = FALSE
)
```

```{r characterization-inspect-results}
characterization_targets <- filter_dqd_results(
  characterization_before,
  check_names = c("withinVisitDates", "plausibleStartBeforeEnd")
)

characterization_targets[
  characterization_targets$numViolatedRows > 0,
  c("checkName", "checkLevel", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows", "thresholdValue")
]
```

## Stage 3 Investigation SQL

### Observations outside the visit window

```{r characterization-observation-sql, results='asis'}
cat("```sql\n", get_issue_sql("characterization_observation_within_visit_dates"), "\n```", sep = "")
```

```{r characterization-find-observation-dates}
characterization_connection <- connect_duckdb(characterization_work_db, read_only = TRUE)
observation_window_failures <- dbGetQuery(
  characterization_connection,
  get_issue_sql("characterization_observation_within_visit_dates")
)
disconnect_duckdb(characterization_connection)

head(observation_window_failures, 20)
```

### Drug exposures with end date before start date

```{r characterization-drug-sql, results='asis'}
cat("```sql\n", get_issue_sql("characterization_drug_start_before_end"), "\n```", sep = "")
```

```{r characterization-find-drug-dates}
characterization_connection <- connect_duckdb(characterization_work_db, read_only = TRUE)
drug_date_failures <- dbGetQuery(
  characterization_connection,
  get_issue_sql("characterization_drug_start_before_end")
)
disconnect_duckdb(characterization_connection)

drug_date_failures
```

## Stage 3 Fixes

```{r characterization-fix-issues}
characterization_connection <- connect_duckdb(characterization_work_db, read_only = FALSE)
apply_characterization_fixes(characterization_connection, stage = "quarto-characterization")
disconnect_duckdb(characterization_connection)
```

## Stage 3 Rerun and Verification

```{r characterization-rerun-dqd}
characterization_after <- run_dqd(
  db_path = characterization_work_db,
  cdm_source_name = "SyntheaTutorialCharacterizationFixed",
  check_severity = "characterization",
  output_json_path = characterization_after_json,
  verbose_mode = TRUE,
  write_to_table = FALSE
)
```

```{r characterization-verify}
characterization_after_targets <- filter_dqd_results(
  characterization_after,
  check_names = c("withinVisitDates", "plausibleStartBeforeEnd")
)

characterization_after_targets[
  characterization_after_targets$numViolatedRows > 0,
  c("checkName", "checkLevel", "cdmTableName", "cdmFieldName", "numViolatedRows", "pctViolatedRows", "thresholdValue")
]
```

::: {.callout-note}
Expected outcome: the tutorial-specific characterization findings should disappear after the rerun.
:::

## Wrap-up and Discussion

This tutorial used prepared snapshots to keep the exercise deterministic while still letting you inspect the failing rows, apply live fixes, and verify the changes with DQD reruns.

From here you can:

- compare the tutorial snapshots against the clean baseline
- open the saved JSON outputs in the DQD dashboard
- adapt the helper functions to create additional teaching scenarios
